---
title: "Your first intent project"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Your first intent project}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`intent` is a project dependency manager for R. It keeps the project contract in
`DESCRIPTION`, uses `renv.lock` as the resolved state, and installs packages
into the project library.

The core workflow is:

```text
DESCRIPTION declares.
intent enforces.
renv executes.
renv.lock records.
```

This article walks through the first project journey: initialize, add a package,
inspect drift, verify the contract, remove a package, and sync after changes.

## Start a project

Create or choose a project directory, then initialize it with an explicit
repository:

```{r}
intent::init(
  path = "analysis-project",
  repos = c(RSPM = "https://packagemanager.posit.co/cran/latest")
)
```

`intent::init()` creates the project files that make the environment
reproducible:

- `DESCRIPTION`
- `renv.lock`
- `.Rprofile`
- `.Renviron`
- `renv/`

It also writes the repository policy to `DESCRIPTION`:

```text
Config/intent/repos/RSPM: https://packagemanager.posit.co/cran/latest
```

That repository name matters. If a lockfile package records `Repository: RSPM`,
the project must declare `Config/intent/repos/RSPM`.

## Add a dependency

Use `add()` for runtime dependencies:

```{r}
intent::add("dplyr", project = "analysis-project")
```

Use `dev = TRUE` for development or optional dependencies:

```{r}
intent::add("testthat", dev = TRUE, project = "analysis-project")
```

Adding a package updates the manifest, installs into the project library, and
snapshots a normalized lockfile.

## Preview before changing state

Mutating commands support dry runs:

```{r}
intent::add("ggplot2", project = "analysis-project", dry_run = TRUE)
```

The returned plan lists what would be installed, what would be written to
`DESCRIPTION`, and whether `renv.lock` would be updated.

## Inspect status

Use `status()` to see drift without changing anything:

```{r}
intent::status(project = "analysis-project")
```

Status reports:

- packages declared in `DESCRIPTION`;
- packages recorded in `renv.lock`;
- packages missing from the project library;
- source policy violations.

The CLI can return the same information as JSON:

```sh
intent status --project analysis-project --json
```

## Verify the project contract

Use `verify()` when you want a stricter check for CI or release preparation:

```{r}
intent::verify(project = "analysis-project")
```

`doctor()` is an alias:

```{r}
intent::doctor(project = "analysis-project")
```

Verification checks that:

- packages declared in `DESCRIPTION` are present in `renv.lock`;
- locked packages are installed in the project library;
- `renv.lock` repositories match `Config/intent/repos/`;
- source policy violations are absent;
- lockfile packages are reachable from declared dependencies or bootstrap tools.

The CLI exits with an error when verification fails, which makes it suitable for
automation:

```sh
intent verify --project analysis-project --json
```

## Remove a dependency

Use `remove()` to delete a direct dependency and reconcile the lockfile:

```{r}
intent::remove("dplyr", project = "analysis-project")
```

The package is removed from `DESCRIPTION`, removed from the project library when
possible, and pruned from the lockfile if it is no longer needed.

## Sync after pulling changes

After editing `DESCRIPTION` manually or pulling changes from Git, use `sync()`:

```{r}
intent::sync(project = "analysis-project")
```

`sync()` installs missing declared packages, prunes lockfile packages no longer
reachable from the manifest, snapshots a normalized lockfile, and restores the
project library.

## A compact daily loop

Most projects settle into a short loop:

```{r}
intent::add("glue")
intent::status()
intent::verify()
intent::sync()
```

The goal is not to replace `renv`. The goal is to make the project contract
explicit, reviewable, and consistent across machines.
